---
title: "Model3_Bayes_Renewable_Energy1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Default Baseline Model
```{r}

# load packages
library(brms)
library(ggplot2)

# set seed
set.seed(42)

# set cores
options(mc.cores = parallel::detectCores())

# set path
data_path <- "data/World_Energy_Consumption.csv"

# set model constants
year_use <- 2018
target <- "renewables_share_energy"
features <- c(
  "gas_share_elec",
  "fossil_elec_per_capita",
  "coal_share_energy",
  "energy_per_gdp",
  "energy_per_capita"
)

```


```{r}

# read data
energy <- read.csv(data_path)

# keep 2018 sovereign countries
energy_2018 <- energy[
  energy$year == year_use &
    !is.na(energy$iso_code) &
    nchar(energy$iso_code) == 3,
]

# keep needed columns
energy_2018 <- energy_2018[, c("country", "iso_code", features, target)]

# drop missing rows
energy_2018 <- na.omit(energy_2018)

# reset row names
rownames(energy_2018) <- NULL

# print checks
cat("year:", year_use, "\n")
cat("countries with complete data:", nrow(energy_2018), "\n")
cat("features:\n")
print(features)
cat("target:", target, "\n")

# preview data
head(energy_2018)

```


```{r}

# check dimensions
dim(energy_2018)

# check country count
length(unique(energy_2018$country))

# check missing values
colSums(is.na(energy_2018))

# check target summary
summary(energy_2018[[target]])
sd(energy_2018[[target]])

```


```{r}

# make scaled copy
energy_scaled <- energy_2018

# scale predictors
energy_scaled[, features] <- scale(energy_2018[, features])

# preview scaled data
head(energy_scaled)

# check scaled predictor means
round(colMeans(energy_scaled[, features]), 6)

# check scaled predictor sds
apply(energy_scaled[, features], 2, sd)

```


```{r}

# define formula
energy_formula <- renewables_share_energy ~
  gas_share_elec +
  fossil_elec_per_capita +
  coal_share_energy +
  energy_per_gdp +
  energy_per_capita

# inspect default priors
energy_prior <- default_prior(
  energy_formula,
  data = energy_scaled,
  family = gaussian()
)

energy_prior

```


```{r}

# fit gaussian bayesian linear regression
energy_brm <- brm(
  energy_formula,
  family = gaussian(),
  data = energy_scaled,
  chains = 4,
  cores = getOption("mc.cores", 1),
  iter = 1000,
  warmup = 180,
  thin = 1,
  prior = energy_prior
)

```


```{r}

# print summary
summary(energy_brm)

# plot chains and posteriors
plot(energy_brm)

# get bayesian r squared
bayes_R2(energy_brm)

```


```{r}

# get posterior predictive draws
post_preds <- posterior_predict(energy_brm)

# get posterior predictive means
energy_2018$bayes_pred_mean <- colMeans(post_preds)

# get residuals
energy_2018$bayes_residual <- energy_2018[[target]] - energy_2018$bayes_pred_mean

# get absolute residuals
energy_2018$abs_bayes_residual <- abs(energy_2018$bayes_residual)

# sort by residual
energy_results <- energy_2018[order(energy_2018$bayes_residual, decreasing = TRUE), ]

# preview top residuals
head(energy_results[, c("country", target, "bayes_pred_mean", "bayes_residual")], 10)

# preview bottom residuals
tail(energy_results[, c("country", target, "bayes_pred_mean", "bayes_residual")], 10)

```


```{r}

# plot posterior predictive means vs actual
plot(
  energy_2018$bayes_pred_mean,
  energy_2018[[target]],
  main = "posterior predictive means vs actual",
  xlab = "posterior predictive mean",
  ylab = "actual renewables share"
)

abline(0, 1, col = "red", lwd = 2)

# check correlation
cor(energy_2018$bayes_pred_mean, energy_2018[[target]])

```


# Adding manual priors
```{r}

# get target mean for intercept prior
target_mean <- mean(energy_2018[[target]])
target_mean

```


```{r}

# define manual priors
manual_prior <- c(
  prior(student_t(3, 14.367, 10), class = "Intercept"),
  prior(normal(0, 5), class = "b", coef = "gas_share_elec"),
  prior(normal(0, 5), class = "b", coef = "fossil_elec_per_capita"),
  prior(normal(0, 5), class = "b", coef = "coal_share_energy"),
  prior(normal(0, 5), class = "b", coef = "energy_per_gdp"),
  prior(normal(0, 5), class = "b", coef = "energy_per_capita"),
  prior(student_t(3, 0, 10), class = "sigma")
)

```


```{r}

# fit manual prior model
energy_brm_manual <- brm(
  energy_formula,
  family = gaussian(),
  data = energy_scaled,
  chains = 4,
  cores = getOption("mc.cores", 1),
  iter = 4000,
  warmup = 720,
  thin = 1,
  prior = manual_prior
)

```


```{r}

# check summary
summary(energy_brm_manual)

# check chains and posteriors
plot(energy_brm_manual)

# check bayesian r squared
bayes_R2(energy_brm_manual)

```


```{r}

# get posterior predictive draws
post_preds_manual <- posterior_predict(energy_brm_manual)

# get posterior predictive means
energy_2018$bayes_pred_mean_manual <- colMeans(post_preds_manual)

# get residuals
energy_2018$bayes_residual_manual <- energy_2018[[target]] - energy_2018$bayes_pred_mean_manual

# get absolute residuals
energy_2018$abs_bayes_residual_manual <- abs(energy_2018$bayes_residual_manual)

# sort by residual
energy_results_manual <- energy_2018[order(energy_2018$bayes_residual_manual, decreasing = TRUE), ]

# preview top residuals
head(
  energy_results_manual[, c("country", target, "bayes_pred_mean_manual", "bayes_residual_manual")], 10)

# preview bottom residuals
tail(energy_results_manual[, c("country", target, "bayes_pred_mean_manual", "bayes_residual_manual")], 10)

```


```{r}

# plot posterior predictive means vs actual
plot(
  energy_2018$bayes_pred_mean_manual,
  energy_2018[[target]],
  main = "manual prior posterior predictive means vs actual",
  xlab = "posterior predictive mean",
  ylab = "actual renewables share"
)

abline(0, 1, col = "red", lwd = 2)

# check correlation
cor(energy_2018$bayes_pred_mean_manual, energy_2018[[target]])

```


```{r}

# compare bayesian r squared
bayes_R2(energy_brm)
bayes_R2(energy_brm_manual)

# compare coefficient summaries
summary(energy_brm)
summary(energy_brm_manual)

```

The default-prior and manual-prior models told the same overall story, which is reassuring: the main coefficient directions were stable, the same broad countries remained over and under-predicted, and the posterior diagnostics improved with the longer manual-prior run. Although the manual-prior model had a slightly lower Bayesian R2 than the default-prior version, it gave a more disciplined Bayesian specification by replacing flat coefficient priors with weakly informative Normal priors and produced cleaner posterior interpretation. In particular, the manual-prior fit reduced uncertainty enough that fossil_elec_per_capita became credibly negative and energy_per_capita became credibly positive, while gas_share_elec and coal_share_energy remained clearly negative and energy_per_gdp remained uncertain. Because the manual-prior specification improves interpretability without changing the overall substantive conclusions, we move forward with it as our primary Bayesian model.

# Cross Validation
```{r}

# set cv constants
k_folds <- 5
cv_iter <- 4000
cv_warmup <- 720
cv_thin <- 1

```


```{r}

# create target tiers
bins <- c(0, 5, 15, 30, 100)
labels <- c("very_low", "low", "medium", "high")

# add tiers
energy_2018$tier <- cut(energy_2018[[target]], breaks = bins, labels = labels, include.lowest = TRUE)

# check tier counts
tier_counts <- table(energy_2018$tier)
tier_counts

```


```{r}

# set seed
set.seed(42)

# initialize fold vector
fold_assignments <- rep(NA_integer_, nrow(energy_2018))

# assign folds within each tier
for (tier_label in levels(energy_2018$tier)) {
  tier_idx <- which(energy_2018$tier == tier_label)
  tier_idx <- sample(tier_idx)
  
  for (i in seq_along(tier_idx)) {
    fold_assignments[tier_idx[i]] <- (i - 1) %% k_folds + 1
  }
}

# store folds
energy_2018$fold <- fold_assignments

# check fold balance
table(energy_2018$tier, energy_2018$fold)

```


```{r}

# define fold fit function
fit_bayes_fold <- function(train_data, test_data, target, features, iter, warmup, thin) {
  
  # make scaled copies
  train_scaled <- train_data
  test_scaled <- test_data
  
  # scale predictors using train fold only
  train_means <- sapply(train_data[, features], mean)
  train_sds <- sapply(train_data[, features], sd)
  
  for (feat in features) {
    train_scaled[[feat]] <- (train_data[[feat]] - train_means[feat]) / train_sds[feat]
    test_scaled[[feat]] <- (test_data[[feat]] - train_means[feat]) / train_sds[feat]
  }
  
  # get train target mean
  train_target_mean <- mean(train_scaled[[target]])
  
  # define manual priors
  fold_prior <- c(
    prior(student_t(3, 14.367, 10), class = "Intercept"),
    prior(normal(0, 5), class = "b", coef = "gas_share_elec"),
    prior(normal(0, 5), class = "b", coef = "fossil_elec_per_capita"),
    prior(normal(0, 5), class = "b", coef = "coal_share_energy"),
    prior(normal(0, 5), class = "b", coef = "energy_per_gdp"),
    prior(normal(0, 5), class = "b", coef = "energy_per_capita"),
    prior(student_t(3, 0, 10), class = "sigma")
)
  
  
  # fit model
  fold_model <- brm(
    energy_formula,
    family = gaussian(),
    data = train_scaled,
    chains = 4,
    cores = getOption("mc.cores", 1),
    iter = iter,
    warmup = warmup,
    thin = thin,
    prior = fold_prior,
    refresh = 0
  )
  
  # get posterior expected draws for test fold
  fold_epred <- posterior_epred(fold_model, newdata = test_scaled)

  # get posterior expected means
  fold_pred_mean <- colMeans(fold_epred)
  
  # return outputs
  list(
    model = fold_model,
    pred_mean = fold_pred_mean,
    y_true = test_scaled[[target]]
  )
}

```


```{r}

# initialize storage
cv_results <- list()
cv_metrics <- data.frame(
  fold = integer(),
  rmse = numeric(),
  r2 = numeric()
)

# initialize oof predictions
energy_2018$cv_bayes_pred <- NA_real_

# run cv
cat("running", k_folds, "fold cross-validation...\n\n")
cat(sprintf("%-6s %-16s %-12s\n", "Fold", "Val RMSE (pp)", "Val R2"))

for (fold in 1:k_folds) {
  
  # split data
  train_data <- energy_2018[energy_2018$fold != fold, ]
  test_data <- energy_2018[energy_2018$fold == fold, ]
  
  # fit fold model
  fold_out <- fit_bayes_fold(
    train_data = train_data,
    test_data = test_data,
    target = target,
    features = features,
    iter = cv_iter,
    warmup = cv_warmup,
    thin = cv_thin
  )
  
  # get predictions
  y_pred <- fold_out$pred_mean
  y_true <- fold_out$y_true
  
  # get fold metrics
  fold_rmse <- sqrt(mean((y_pred - y_true)^2))
  fold_r2 <- 1 - sum((y_true - y_pred)^2) / sum((y_true - mean(y_true))^2)
  
  # store metrics
  cv_metrics <- rbind(
    cv_metrics,
    data.frame(
      fold = fold,
      rmse = fold_rmse,
      r2 = fold_r2
    )
  )
  
  # store oof predictions
  energy_2018$cv_bayes_pred[energy_2018$fold == fold] <- y_pred
  
  # store model object if needed
  cv_results[[paste0("fold_", fold)]] <- fold_out
  
  # print fold summary
  cat(sprintf("%-6d %-16.3f %-12.4f\n", fold, fold_rmse, fold_r2))
}

cat(sprintf("%-6s %-16.3f %-12.4f\n", "Mean", mean(cv_metrics$rmse), mean(cv_metrics$r2)))
cat(sprintf("%-6s %-16.3f %-12.4f\n", "Std", sd(cv_metrics$rmse), sd(cv_metrics$r2)))

```


```{r}

# print final summary
cat("cross-validated rmse:", round(mean(cv_metrics$rmse), 2), "+/-", round(sd(cv_metrics$rmse), 2), "percentage points\n")
cat("cross-validated r2:", round(mean(cv_metrics$r2), 3), "+/-", round(sd(cv_metrics$r2), 3))

```


```{r}

# get oof residuals
energy_2018$cv_bayes_residual <- energy_2018[[target]] - energy_2018$cv_bayes_pred

# get absolute residuals
energy_2018$cv_abs_bayes_residual <- abs(energy_2018$cv_bayes_residual)

# sort results
cv_energy_results <- energy_2018[order(energy_2018$cv_bayes_residual, decreasing = TRUE), ]

# preview top residuals
head(
  cv_energy_results[, c("country", target, "cv_bayes_pred", "cv_bayes_residual")],
  10
)

# preview bottom residuals
tail(
  cv_energy_results[, c("country", target, "cv_bayes_pred", "cv_bayes_residual")],
  10
)

```


```{r}

# plot cv posterior predictive means vs actual
plot(
  energy_2018$cv_bayes_pred,
  energy_2018[[target]],
  main = "cv posterior predictive means vs actual",
  xlab = "cv posterior predictive mean",
  ylab = "actual renewables share"
)

# add reference line
abline(0, 1, col = "red", lwd = 2)

# check correlation
cor(energy_2018$cv_bayes_pred, energy_2018[[target]])

```
We move forward with this Bayesian model because it gives us strong out-of-sample performance while still being easy to interpret. Its cross-validated results are essentially the same as the MLP’s, which suggests the current five predictors already capture most of the predictive signal available in this setup. Because the Bayesian model matches the MLP so closely without adding extra functional complexity, we keep the simpler specification as our primary Bayesian model.



```{r}

# print fold metrics
cv_metrics

# print mean and sd
data.frame(
  metric = c("rmse", "r2"),
  mean = c(mean(cv_metrics$rmse), mean(cv_metrics$r2)),
  sd = c(sd(cv_metrics$rmse), sd(cv_metrics$r2))
)

```


```{r}

# set layout
par(mfrow = c(1, 2))

# plot rmse by fold
barplot(
  cv_metrics$rmse,
  names.arg = cv_metrics$fold,
  main = "cv rmse by fold",
  xlab = "fold",
  ylab = "rmse (percentage points)",
  col = "steelblue"
)

abline(h = mean(cv_metrics$rmse), col = "red", lwd = 2, lty = 2)

# plot r2 by fold
barplot(
  cv_metrics$r2,
  names.arg = cv_metrics$fold,
  main = "cv r2 by fold",
  xlab = "fold",
  ylab = "r2",
  col = "steelblue"
)

abline(h = mean(cv_metrics$r2), col = "red", lwd = 2, lty = 2)

# reset layout
par(mfrow = c(1, 1))

```


```{r}

# get top and bottom residuals
top10_cv <- head(cv_energy_results, 10)
bot10_cv <- tail(cv_energy_results, 10)

# combine and sort
cv_bar_data <- rbind(top10_cv, bot10_cv)
cv_bar_data <- cv_bar_data[order(cv_bar_data$cv_bayes_residual), ]

# set colors
bar_colors <- ifelse(cv_bar_data$cv_bayes_residual > 0, "seagreen", "tomato")

# plot residual bar chart
barplot(
  cv_bar_data$cv_bayes_residual,
  names.arg = cv_bar_data$country,
  horiz = TRUE,
  las = 1,
  col = bar_colors,
  main = "top over- and under-performers",
  xlab = "actual - predicted (percentage points)"
)

abline(v = 0, lwd = 1)

```


```{r}

# plot residual distribution
hist(
  energy_2018$cv_bayes_residual,
  breaks = 18,
  col = "steelblue",
  main = "cv residual distribution",
  xlab = "residual (percentage points)"
)

abline(v = 0, col = "red", lwd = 2, lty = 2)

```


```{r}

# get posterior summaries
coef_summary <- posterior_summary(
  energy_brm_manual,
  pars = c(
    "b_gas_share_elec",
    "b_fossil_elec_per_capita",
    "b_coal_share_energy",
    "b_energy_per_gdp",
    "b_energy_per_capita"
  ),
  probs = c(0.025, 0.975)
)

# make data frame
coef_df <- data.frame(
  term = c(
    "gas_share_elec",
    "fossil_elec_per_capita",
    "coal_share_energy",
    "energy_per_gdp",
    "energy_per_capita"
  ),
  estimate = coef_summary[, "Estimate"],
  lower = coef_summary[, "Q2.5"],
  upper = coef_summary[, "Q97.5"]
)

# order terms
coef_df <- coef_df[order(coef_df$estimate), ]

# plot intervals
plot(
  coef_df$estimate,
  seq_along(coef_df$term),
  xlim = range(c(coef_df$lower, coef_df$upper)),
  yaxt = "n",
  ylab = "",
  xlab = "posterior estimate",
  main = "coefficient credible intervals",
  pch = 19
)

axis(2, at = seq_along(coef_df$term), labels = coef_df$term, las = 1)

segments(
  x0 = coef_df$lower,
  y0 = seq_along(coef_df$term),
  x1 = coef_df$upper,
  y1 = seq_along(coef_df$term),
  lwd = 2
)

abline(v = 0, col = "red", lwd = 2, lty = 2)

```


```{r}

# get posterior draws for coefficients
post_draws <- as.data.frame(energy_brm_manual)

# build coefficient table
coef_table <- data.frame(
  term = c(
    "gas_share_elec",
    "fossil_elec_per_capita",
    "coal_share_energy",
    "energy_per_gdp",
    "energy_per_capita"
  ),
  estimate = c(
    mean(post_draws$b_gas_share_elec),
    mean(post_draws$b_fossil_elec_per_capita),
    mean(post_draws$b_coal_share_energy),
    mean(post_draws$b_energy_per_gdp),
    mean(post_draws$b_energy_per_capita)
  ),
  lower_95 = c(
    quantile(post_draws$b_gas_share_elec, 0.025),
    quantile(post_draws$b_fossil_elec_per_capita, 0.025),
    quantile(post_draws$b_coal_share_energy, 0.025),
    quantile(post_draws$b_energy_per_gdp, 0.025),
    quantile(post_draws$b_energy_per_capita, 0.025)
  ),
  upper_95 = c(
    quantile(post_draws$b_gas_share_elec, 0.975),
    quantile(post_draws$b_fossil_elec_per_capita, 0.975),
    quantile(post_draws$b_coal_share_energy, 0.975),
    quantile(post_draws$b_energy_per_gdp, 0.975),
    quantile(post_draws$b_energy_per_capita, 0.975)
  ),
  prob_positive = c(
    mean(post_draws$b_gas_share_elec > 0),
    mean(post_draws$b_fossil_elec_per_capita > 0),
    mean(post_draws$b_coal_share_energy > 0),
    mean(post_draws$b_energy_per_gdp > 0),
    mean(post_draws$b_energy_per_capita > 0)
  ),
  prob_negative = c(
    mean(post_draws$b_gas_share_elec < 0),
    mean(post_draws$b_fossil_elec_per_capita < 0),
    mean(post_draws$b_coal_share_energy < 0),
    mean(post_draws$b_energy_per_gdp < 0),
    mean(post_draws$b_energy_per_capita < 0)
  )
)

coef_table

```


```{r}

# get posterior expected draws for all countries from full manual-prior model
country_epred <- posterior_epred(energy_brm_manual, newdata = energy_scaled)

# build prediction interval table
country_pred_table <- data.frame(
  country = energy_2018$country,
  actual = energy_2018[[target]],
  pred_mean = colMeans(country_epred),
  pred_lower_95 = apply(country_epred, 2, quantile, 0.025),
  pred_upper_95 = apply(country_epred, 2, quantile, 0.975)
)

# show selected countries
country_pred_table[country_pred_table$country %in% c(
  "Norway", "Brazil", "New Zealand", "Luxembourg", "France", "Hong Kong"
), ]

```


```{r}

# get posterior predictive draws for all countries
country_post_pred <- posterior_predict(energy_brm_manual, newdata = energy_scaled)

# residual draws
residual_draws <- sweep(country_post_pred, 2, energy_2018[[target]], FUN = "-")
residual_draws <- -residual_draws

# build residual probability table
country_resid_table <- data.frame(
  country = energy_2018$country,
  actual = energy_2018[[target]],
  pred_mean = colMeans(country_post_pred),
  resid_mean = colMeans(residual_draws),
  resid_lower_95 = apply(residual_draws, 2, quantile, 0.025),
  resid_upper_95 = apply(residual_draws, 2, quantile, 0.975),
  prob_overperform = colMeans(residual_draws > 0),
  prob_underperform = colMeans(residual_draws < 0)
)

# sort by posterior mean residual
country_resid_table <- country_resid_table[order(country_resid_table$resid_mean, decreasing = TRUE), ]

# preview top and bottom
head(country_resid_table, 10)
tail(country_resid_table, 10)

```


```{r}

# select top and bottom countries by residual mean
plot_data <- rbind(
  head(country_resid_table, 5),
  tail(country_resid_table, 5)
)

# order for plotting
plot_data <- plot_data[order(plot_data$prob_overperform), ]

# plot posterior probability of over-performance
barplot(
  plot_data$prob_overperform,
  names.arg = plot_data$country,
  horiz = TRUE,
  las = 1,
  xlim = c(0, 1),
  col = "steelblue",
  main = "posterior probability of over-performance",
  xlab = "probability actual > predicted"
)

```


```{r}

# build output table to match mlp csv structure
bayes_results_export <- energy_2018[, c("country", "iso_code", target, "tier")]

# use cv predictions as oof values
bayes_results_export$oof_predicted <- energy_2018$cv_bayes_pred
bayes_results_export$oof_residual <- energy_2018$cv_bayes_residual
bayes_results_export$abs_residual <- abs(energy_2018$cv_bayes_residual)

# use full manual-prior model predictions as final values
bayes_results_export$final_predicted <- energy_2018$bayes_pred_mean_manual
bayes_results_export$final_residual <- energy_2018$bayes_residual_manual

# order columns exactly like mlp output
bayes_results_export <- bayes_results_export[, c(
  "country",
  "iso_code",
  target,
  "oof_predicted",
  "oof_residual",
  "final_predicted",
  "final_residual",
  "abs_residual",
  "tier"
)]

# save csv
write.csv(
  bayes_results_export,
  "bayes_country_predictions.csv",
  row.names = FALSE
)

```